Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-06 Manuscript/manuscript.tex
Pavel's corrections to the MS.

\begin{abstract}
  % insert abstract here
  The binding affinities and stoichimetries of Na$^+$ and Ca$^{2+}$ ions to phospholipid bilayers have not been
  consistently determined with different experimental and theoretical methods
  despite their significance and biological relevance in cellular membranes. 
  The ion binding details could be resolved with classical molecular dynamics (MD) simulations,
  however, the accuracy of the available lipid models has not been sufficient.
  In this work we show that the binding details of Na$^+$ and Ca$^{2+}$ ions to
  1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
  can be accurately described with a MD simulation model having implicitly included electronic polarization.
  This is demonstrated by applying the electronic continuum correction (ECC) to a
  state of the art lipid model for MD simulations of POPC
  lipid bilayer. The introduced ECC-lipids model reproduces the experimentally measured structural parameters for the ion-free membrane,
  the response of lipid headgroup to the bound positive charge, and the binding affinities of
  Na$^+$ and Ca$^{2+}$ ions. The imperceptible binding of Na$^+$ ions to POPC bilayer
  and interactions of Ca$^{2+}$ mainly with phoshpate oxygens in the ECC-lipids model give
  support to the early intepretations of the experimental spectroscopic data.
  On the other hand, Ca$^{2+}$ ions form complexes with 1-3 lipid molecules with almost
  equal probabilities, suggesting more complicated binding stoichimetry than 
  the ternary complex binding model used to interpret the NMR data.
  The results in this work pave the way for MD simulations of complex biochemical systems
  with realistic electrostatic interactions in the vicinity of cellular membranes.
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as signal propagation in neurons and vesicle fusion.
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. Especially the zwitterionic phosphocholine (PC) lipid bilayers are
used to understand the role of ions in complex biological
systems~\cite{scherer87,seelig90,cevc90}.

Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Interpretations of non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared spectroscopy suggest that Na$^+$ ions exhibit
negligible binding to PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ is observed to specifically
bind to phosphate groups of two lipid molecules~\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Atomistic resolution molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Furthermore, interactions of calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

Recent work published by the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve the apparent controversies. A direct comparison of
ion binding affinities to PC bilayers was presented between simulations and
experiments by using the electrometer concept, which is based on the experimental NMR data
for the lipid headgroup order parameters~\cite{seelig87}.
Using massive amounts of data collected by Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for the detailed interpretation of the cation
interactions with PC lipid bilayers~\cite{catte16}.

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved with an implicit inclusion of the electronic polarizability
in the polar region of phospholipids,
which we apply according to the 
electronic continuum correction (ECC)~\cite{leontyev11}.
Such an approach has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
We use the the Lipid14 model \cite{dickson14} as a starting point,
because it provided one of the best available descriptions of cation binding \cite{catte16}.
The developed ECC-lipids model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly in the description of the membrane binding affinities to cations and induced
structural effects of sodium and calcium ions.

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of polarizability in standard MD simulation
force fields has been considered an issue since the early days of
lipid bilayer simulations.
In this work, we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{lucas12,chowdhary13},
including the electronic part of polarizability in lipid
bilayer simulations implicitly through 
the electronic continuum correction (ECC)~\cite{leontyev11}. 
Technically, it is similar to the phenomenological charge-scaling applied in 
earlier studies \cite{jonsson86,egberts94}.
\todo{We should also cite papers where empirical scaling was used ionic liquids DOI: 10.1002/anie.201308760}
However, the present concept of ECC is
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.

According to ECC, electronic
polarizability can be implicitly included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. 
Following the Coulomb's law, such a dielectric continuum can be 
easily included in a standard MD simulation by
the scaling of charges,
\begin{equation}
  Q^{ECC} = f_q \cdot Q ,
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
Given the  high frequency dielectric constant as measured in water (corresponding to the square of the refraction index) of  $\epsilon _{el} = 1.78$,  
the scaling factor for aqueous ions is $f_q = 0.75$ \cite{some_orig_source, leontyev11}. This scaling factor has been
successfully used to improve the performance of force field for ions in solution \cite{kohagen14,kohagen16,??},
which then agree quantitatively with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014}.

While the scaling factor of $f_q = 0.75$ for ions in water improves their description and is physically well justified within the ECC theory \cite{Leontyev2011},
it is not a priori  clear whether the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of an atom or molecules, atomic partial charges within molecules are not physical observables.
There are thus several schemes for the assignment of partial charges for biomolecules~\cite{Hu2007}.
Currently, the most commonly employed scheme is the restrained electrostatic potential (RESP) method~\cite{RESP_paper,Singh1984}.
By construction, partial charges currently implemented in force fields may already 
include  to some extend the solvent electronic polarizability effects, i.e.,
the RESP charges are subsequently modified to fit certain experimental observables \todo{This needs a citation - Joe please pick 1-2 papers}. 
Thus, a consistent application of the ECC scaling factor, $f_q$, to the molecular partial charges
included in the available force fields does not necessarily have to follow the above relation $f_q = \epsilon _{el} ^{-1/2}$, 
but instead it lies between 0.75 i.e., (no electronic polarizability included in the original partial charges)
and $1$ (i.e., electronic polarizability fully included in the original partial charges).
It is important to note that even though the lipid bilayer interface has a large gradient in
the total dielectric constant, the electronic part of polarizability can be considered to be constant
as the measurements of high frequency dielectric constant gives values around 2 
for almost any biologically relevant material \cite{some_original_work, leontyev11}.

In this work, we employ the ECC to develop a classical MD simulation model for
a POPC lipid that accurately describes the binding of sodium and calcium ions to a lipid bilayer.
%This is a biologically highly relevant membrane feature, which is
%poorly reproduced by currently available models. This can affect not only on membrane properties
%in the presence of ions but also modulate interactions with charged moieties at the interface.
The quality of the simulation model is compared 
to the experimental NMR data used to measure the ion binding
affinity~\cite{akutsu81,altenbach84,scherer89}, 
as discussed in detail in Ref.~\citenum{catte16}.
The Lipid14~\cite{dickson14} force field parameters (available in Gromacs format from Ref. \citenum{lipid14files})
were used as a starting point, because they provide one of the most
realistic descriptions of the head group order parameter response to ions 
among the available lipid models (see Figs.~2~and~5 in Ref.~\citenum{catte16}).
Moreover, the Lipid14 model has a relatively realistic glycerol backbone and
head group structure compareable to other state of the art lipid models~\cite{botan15}.

We applied the ECC correction to the Lipid14 model of POPC as a scaling of partial charges of the head group,
glycerol backbone, and carbonyl regions, which are the most polar parts in phospholipids
being expected to have the largest contribution to the cation binding.
We do not modify the hydrocarbon chain parameters, 
as they do not come in direct contact with salt ions, 
and as they are already highly optimized to provide a good description of the
hydrophobic part of lipid bilayers \cite{dickson14,ollila16}.
This is in contrasts with the glycerol backbone and the head group regions, 
which call for improvements in all available lipid models~\cite{botan15}.

The values within the above discussed parameter space for the scaling factor, $f_q~\in~(0.75, 1.0)$,
were explored to find the optimal value to reproduce the experimental ion binding affinity.
We observed that the scaling
of the partial charges reduced the ion binding affinity and the 
related head group order parameter response in general.
We compared the results with different scaling factors to the experimental 
NMR data~\cite{akutsu81,altenbach84,scherer89} finding the optimal 
value $f_q = 0.8$,
which is only slighly higher than the scaling factor for the ions in water, 0.75.

Although the scaling of the partial charges improved 
the head group order parameter response and ion binding affinity,
it also deteriorated some membrane properties; 
namely the area per lipid decreased below the experimental value. 
After only scaling the partial charges of the head group, the glycerol
backbone and carbonyls with a factor of 0.8, 
the area per lipid in the simulation drops to approximately 60~\AA$^2$,
which is significantly smaller than the experimental value, 64.3~\AA$^2$ \cite{kucerka11}, 
as well as the value from the original Lipid14 model, $65.6 \pm 0.5$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid is found to arise from a reduced hydration of the lipid head group region
as a consequence of the lower polarity after scaling charges. 
This effect can be compensated
by reducing the effective radii of atoms with the scaled charges.
This is in practice done by changing the parameters $\sigma$ in the Lennard-Jones potential 
in a similar way as was previously done for the ECC-ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
After reducing the $\sigma$ parameters of affected atoms with a scaling factor $f_\sigma = 0.89$, 
the area per molecule is restored
back close to the experimental value (Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding in lipid bilayers was compared between experiments and simulations 
using the lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}, which
is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in a PC lipid head group
(Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of charge bound per lipid~\cite{seelig87}.
The change of the order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

The concept can be used to compare the ion binding affinity to lipid 
bilayers between MD simulations and  NMR experiments, because the order
parameters can be accurately determined from both techniques~\cite{catte16,ollila16}.
The order parameters for all C-H bonds in lipid molecules, including
$\alpha$ and $\beta$ segments in head group, can be accurately measured
using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the
order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and the average is taken over all sampled configurations \cite{ollila16}.

The relation between the amount of the bound charge per lipid,  $X^\pm$, and
the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$,
is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for the binding of
Ca$^{2+}$ to POPC bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because the head group order parameters
decrease with a bound positive charge and increase with a bound negative
charge~\cite{ollila16,catte16}. This can be rationalized as a change
of the lipid head group dipole tilt towards water phase with the bound positive
charge and {\it vice versa} with the negative charge~\cite{seelig87}.

The measured order parameter change depends on both the head group response 
to the bound charge, and the amount of bound charge 
(i.e. $m_i$ and $X^\pm$ in Eq.~\ref{OPchangeEQ}, respectively). 
The empirical factor $m_i$ has to be
well quantified before the electrometer concept can used to 
analyze the binding affinities.
This was done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response to the bound charge in simulations,
we use the experimental data for dihexadecyldimethylammoniumbromide %(2C$_{16}$$^+$N2C$_1$Br$^-$)
mixed within a POPC bilayer~\cite{scherer89}. Dihexadecyldimethylammonium\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
The concentrations of salts are reported in two different ways
in the NMR measurements of head group order parameters \cite{akutsu81, altenbach84}.
The early work on the NMR electrometer concept 
reports the salt concentrations in water before solvating
the lipids~\cite{akutsu81}. 
The later study uses the atomic absorption
spectroscopy, and reports the salt concentration in the supernatant after the
solvation of lipids~\cite{altenbach84}.
In this work, we focus on POPC for which the latter definition was used~\cite{altenbach84}.
The salt concentration in the aqueous bulk region was calculated from 
the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study~\cite{catte16},
the ion concentrations were calculated in water before solvating the lipids 
as in the earlier experiments~\cite{akutsu81}.
Although there are measurable differences between these two
definitions of concentrations for CaCl$_2$ systems,
they do not affect the qualitative conclusions in this or in the previous work~\cite{catte16}.

>>>>

\begin{abstract}
  % insert abstract here
  The binding affinities and stoichimetries of Na$^+$ and Ca$^{2+}$ ions to phospholipid bilayers have not been
  consistently determined with experimental and computational methods,
  despite their significance and biological relevance in cellular membranes. 
  The details concerning ion binding could be resolved with classical molecular dynamics (MD) simulations,
  however, the accuracy of the available lipid models has not been sufficient.
  In this work we show that the binding details of Na$^+$ and Ca$^{2+}$ ions to the
  1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
  can be accurately described with a MD simulation model implicitly including electronic polarization.
  This is demonstrated by applying the electronic continuum correction (ECC) to a
  state of the art lipid model for MD simulations of the POPC
  lipid bilayer. The newly introduced ECC-lipids model reproduces the experimentally measured structural parameters for the ion-free membrane,
  the response of lipid headgroup to a bound positive charge, and the binding affinities of
  Na$^+$ and Ca$^{2+}$ ions. The observed negligible binding of Na$^+$ ions to POPC bilayer
  and stronger interactions of Ca$^{2+}$ primarily with phoshpate oxygens give
  support to the early intepretations of the experimental spectroscopic data.
  At the same time, Ca$^{2+}$ ions form complexes with 1-3 lipid molecules with almost
  equal probabilities, suggesting a more complicated binding stoichimetry than 
  the ternary complex binding model used to interpret the NMR data.
  The results of this work pave the way to accurate MD simulations of complex biochemical systems
  with realistic electrostatic interactions in the vicinity of cellular membranes.
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in important biological processes,
such as signal propagation in neurons and vesicle fusion.
Since direct measurements of ion-membrane interactions in biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. In particular, the zwitterionic phosphocholine (PC) lipid bilayers are
used as entry level models, helping to understand the role of ions in complex biological
systems~\cite{scherer87,seelig90,cevc90}.

Interactions of biologically relevant cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers have been widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Interpretations of non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), as well as scattering
and infrared spectroscopies, suggest that Na$^+$ ions exhibit only
negligible binding to PC lipid bilayers wat submolar concentrations, while Ca$^{2+}$ is observed to specifically
bind to phosphate groups of two lipid molecules~\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Atomistic resolution molecular dynamics (MD) simulation models, however, typically predict a
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, several experiments have also been interpreted to support the predictions from
earlier MD simulations~\cite{bockmann03,vacha09a}. 
Furthermore, interactions of calcium ions with up to 3-4 lipids,
including not only interactions with phosphates but also with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

Recent studies within the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve these apparent controversies. A direct comparison of
ion binding affinities to PC bilayers between simulations and
experiments was done using the electrometer concept, which is based on experimental NMR data
for the lipid headgroup order parameters~\cite{seelig87}.
Analyzing massive amounts of data collected by the Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for a detailed interpretation of the
interactions of cations with PC lipid bilayers~\cite{catte16}.

@@

In this work, we show that the cation binding behavior
in MD simulations of the 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved via an implicit inclusion of the electronic polarizability
in the polar region of phospholipids,
which we apply using to the 
electronic continuum correction (ECC)~\cite{leontyev11}.
Such an approach has been previously shown to improve the behaviour of
of ions in bulk water simulations ~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
We employ here the the Lipid14 model \cite{dickson14} as a starting point,
because it provided one of the best available descriptions of cation binding \cite{catte16}.
The newly developed ECC-lipids model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly in the description of the membrane binding affinities to cations and induced
structural effects of sodium and calcium ions.

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in standard MD simulation
force fields has been considered a serious issue since the early days of
lipid bilayer simulations.
In this work, we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{lucas12,chowdhary13} by
including the electronic part of polarizability in lipid
bilayer simulations implicitly via 
the electronic continuum correction (ECC)~\cite{leontyev11}. 
Technically, this is similar to the phenomenological charge-scaling applied in 
earlier studies of aqueous solutions or ionic liquids \cite{jonsson86,egberts94}.
\todo{We should also cite papers where empirical scaling was used ionic liquids DOI: 10.1002/anie.201308760}
However, the present concept of ECC is
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.

According to ECC, the electronic
polarizability can be implicitly included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the medium \cite{leontyev11}. 
Following the Coulomb's law, such a dielectric continuum can be 
directly included in a standard MD simulation by
scaling of charges,
\begin{equation}
  Q^{ECC} = f_q \cdot Q ,
\end{equation}
with a constant scaling factor of $f_q = \epsilon _{el} ^{-1/2}$ 
. 
Given the  high frequency dielectric constant as measured in water (corresponding to the square of the refraction index) of  $\epsilon _{el} = 1.78$,  
the scaling factor for aqueous ions is $f_q = 0.75$ \cite{some_orig_source, leontyev11}. This scaling factor has been
successfully used to improve the performance of force field simulations of ions in aqueous solution \cite{kohagen14,kohagen16,??},
which then agree quantitatively with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014}.

While the scaling factor of $f_q = 0.75$ for ions in water improves their description and is physically well justified within the ECC theory \cite{Leontyev2011},
it is not a priori  clear whether the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of an ion, atomic partial charges within molecules are not physical observables.
Several schemes exist for the assignment of partial charges for biomolecules~\cite{Hu2007}.
Currently, the most commonly employed scheme is the restrained electrostatic potential (RESP) method~\cite{RESP_paper,Singh1984}.
By construction, partial charges currently implemented in force fields may already 
include  to some extend the electronic polarizability effects of the solvent, i.g.,
the RESP charges may be subsequently modified to fit certain experimental observables \todo{This needs a citation - Joe please pick 1-2 papers}. 
Thus, a consistent application of the ECC scaling factor, $f_q$, to the partial charges in molecules as
included in the available force fields does not necessarily have to follow the above relation $f_q = \epsilon _{el} ^{-1/2}$. 
Instead, a consistent scaling factor would lie between the value of 0.75 i.e., (no electronic polarizability included in the original partial charges)
and $1$ (i.e., electronic polarizability fully included in the original partial charges).
It is important to note that even though the lipid bilayer interface has a large gradient in
the total dielectric constant, the electronic part of polarizability can be considered to be almost constant
as measurements give values around 2 
for almost any biologically relevant systems \cite{some_original_work, leontyev11}.

In this work, we employ ECC to develop a force field model for
the POPC lipid that accurately describes the binding of sodium and calcium ions to the corresponding lipid bilayer.
%This is a biologically highly relevant membrane feature, which is
%poorly reproduced by currently available models. This can affect not only on membrane properties
%in the presence of ions but also modulate interactions with charged moieties at the interface.
The quality of the simulation model is judged by comparison 
to experimental NMR data reporting on the ion binding
affinity~\cite{akutsu81,altenbach84,scherer89}, 
as discussed in detail in Ref.~\citenum{catte16}.
The Lipid14~\cite{dickson14} force field parameters (available in a Gromacs format from Ref. \citenum{lipid14files})
were used as a starting point, as they provide one of the most
realistic descriptions of the head group order parameter response to ions 
among the available lipid models (see Figs.~2~and~5 in Ref.~\citenum{catte16}).
Moreover, the Lipid14 model has a relatively realistic glycerol backbone and
head group structure comparable to other state of the art lipid models~\cite{botan15}.

We applied the ECC correction to the Lipid14 model of POPC by scaling of partial charges of the head group,
glycerol backbone, and carbonyl regions, which are the polar parts in phospholipids,
expected to have a contribution to the cation binding.
We did not modify the hydrocarbon chain parameters, 
as they do not come in direct contact with ions from the solution 
and as they are already highly optimized to provide a good description of the
hydrophobic part of lipid bilayers \cite{dickson14,ollila16}.
The latter is in contrasts with the situation for the glycerol backbone and the head group regions, 
which calls for improvements in available lipid models~\cite{botan15}.

The above discussed interval for the scaling factor, $f_q~\in~(0.75, 1.0)$,
was explored to find the optimal value reproducing the experimental ion binding affinities.
We first observed that the scaling
of the partial charges reduced the ion binding affinity and the 
related head group order parameter response in general.
We then compared the results with different scaling factors to the experimental 
NMR data~\cite{akutsu81,altenbach84,scherer89} finding the optimal 
value $f_q = 0.8$,
which is only slighly higher than the original scaling factor for the ions in water of 0.75.

Although scaling of the partial charges improved 
the head group order parameter response and ion binding affinity,
it at the same time deteriorated certain membrane properties; 
namely the area per lipid decreased below the experimental value. 
After directly scaling the partial charges of the head group, the glycerol
backbone, and carbonyls with a factor of 0.8, 
the area per lipid in the simulation drops approximately to 60~\AA$^2$,
which is smaller than the experimental value of 64.3~\AA$^2$ \cite{kucerka11}, 
as well as the value from the original Lipid14 model of $65.6 \pm 0.5$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid is found to arise from a reduced hydration of the lipid head group region
due a reduced head group polarity after scaling charges. 
This effect is compensated
by reducing the effective radii of atoms with the scaled charges.
This is in practice done by changing the parameters $\sigma$ in the Lennard-Jones potential 
in a similar way as was done previously for the ECC ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
After reducing the $\sigma$ parameters of affected atoms by a factor of $f_\sigma = 0.89$, 
the area per molecule is restored
back, being very close to the experimental value (Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding in lipid bilayers was compared between experiments and simulations 
using the lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}, which
is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in a PC lipid head group
(Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of charge bound per lipid~\cite{seelig87}.
The change of the order parameters measured with varying aqueous
ion concentrations can be then related to the amount of bound ions.

The electrometer concept can be used to compare the ion binding affinity to lipid 
bilayers between MD simulations and  NMR experiments, because the order
parameters can be accurately determined from both techniques~\cite{catte16,ollila16}.
The order parameters for all C-H bonds in lipid molecules, including
$\alpha$ and $\beta$ segments in head group, can be accurately measured
using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the
order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and the average is taken over all sampled configurations \cite{ollila16}.

The relation between the amount of the bound charge per lipid,  $X^\pm$, and
the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$,
is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for the binding of
Ca$^{2+}$ to POPC bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because the head group order parameters
decrease with a bound positive charge and increase with a bound negative
charge~\cite{ollila16,catte16}. This can be rationalized in terms of a change
of the lipid head group dipole tilt towards water phase with the bound positive
charge and {\it vice versa} with the negative charge~\cite{seelig87}.

The measured change of the order parameter depends both on the head group response 
to the bound charge and on the amount of the bound charge 
(i.e. $m_i$ and $X^\pm$ in Eq.~\ref{OPchangeEQ}, respectively). 
The empirical factor $m_i$ has to be
properly quantified before the electrometer concept can used to 
analyze the binding affinities.
This has been done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the response of the head group order parameter to the bound charge in simulations,
we use experimental data for a strong cationic surfactant dihexadecyldimethylammoniumbromide %(2C$_{16}$$^+$N2C$_1$Br$^-$)
mixed within a POPC bilayer~\cite{scherer89}. Dihexadecyldimethylammonium\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cationic surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and its molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
The concentrations of salts are reported in two different ways
in the NMR measurements of head group order parameters \cite{akutsu81, altenbach84}.
The early work on the NMR electrometer concept 
reports the salt concentrations in water before solvating
the lipids~\cite{akutsu81}. 
The later study uses the atomic absorption
spectroscopy, and reports the salt concentration in the supernatant after the
solvation of lipids~\cite{altenbach84}.
In this work, we focus on POPC for which the latter definition was used~\cite{altenbach84}.
The salt concentration in the aqueous bulk region was taken as that at 
the farthest point from the lipid bilayer in the aqueous phase. 
Note that in the previous study~\cite{catte16},
the ion concentrations were calculated in water before solvating the lipids, 
as in the earlier experiments~\cite{akutsu81}.
Although there are noticable differences between these two
definitions for concentrations of the CaCl$_2$ systems,
they do not affect significantly conclusions drawn in this work or in the previous one~\cite{catte16}.

To quantify the ion binding affinities to a lipid bilayer, we
calculate the relative surface excess of ions with respect to water,~$\Gamma_i^w$~\cite{chattorajBOOK}.
Such a quantity compares the adsorption of ions to the adsorption of water molecules 
at the interface without the necessity of defining a Gibbs dividing plane 
between the membrane interior and the water bulk region. 
In our simulations we only assume that the interface is located between the ion-free hydrophobic interior
of the lipid bilayer and the aqueous region far from the membrane.
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
Such a setup and the above definition of bulk ion concentration provide a simple relation for the normalized surface excess $\Gamma_i^w$ for simulations of lipid bilayers, 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) .
\end{equation}
Here, $n_w$ and $n_i$ are the total numbers of water molecules and ions in the system,
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase,
and $A_b$ is the area of the unit cell in the membrane plane.
The total area of the interface is then twice the area of the membrane, i.e., $2A_b$,
because the bilayer has an interface at each of the two leaflets.
